Tilting at Algorithmic Windmills

### 2013-05-05 Tilting/paper.tex
ENH:  Adding tilting paper.

\title{Multivariate Neuroanalysis with ANTsR:  Application to Supervised Brain Segmentation}

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{K.~L.~Shrinidhi}
\author[label2]{Jeffrey T.~Duda}
\author[label2]{Philip A.~Cook}
\author[label1]{Christopher Durst}
\author[label1]{James C.~Gee}
\author[label1]{Murray C.~Grossman}
\author[label1]{Max Wintermark}
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, University of Pennsylvania,
                Philadelphia, PA}

@@

\begin{abstract} 
Improvements in image acquisition and increasingly sophisticated statistical techniques 
for neuroanalysis have spurred the development of advanced computational frameworks for studying the
brain---many of which
have been made publicly available.  One such popular toolkit is the open source Advanced
Normalization Tools (ANTs) which contains a suite of well-vetted processing tools for
core data transformation tasks such as image registration, image segmentation, 
inhomogeneity field correction and template building.  However, despite the numerous
solutions afforded by such tools, neuroscience (and data analysis in general) requires
a statistical platform for making inferences with respect to given hypotheses once the 
requisite data transformations have been performed.  In this paper, we describe the
coupling of ANTs with the widely-used R language---an open source environment for 
statistical processing and data visualization---which we denote as ANTsR.  One of the
many benefits from such an integration is the accessibility to advanced statistical 
and machine learning techniques.  To showcase the flexibility and power of ANTsR, we 
apply the combination of one such technique, Random Forests, and ANTs processing 
tools to the difficult problem of brain tumor segmentation.  This includes evaluation on 
the public data set from the MICCAI 2012 BRATS challenge consisting of both real and simulated
data.  To facilitate reproducibility, all scripts and data used in this paper are 
publicly available for download.  
\end{abstract}

\begin{keyword}
advanced normalization tools \sep BRATS \sep brain tumor segmentation \sep R project \sep random forests
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

O

\section{Materials and Methods}

\subsection{ANTsR:  An ANTs/R Interface}

\subsubsection{Installation}

The ANTsR package is publicly available on the github project hosting service.%
\footnote{
https://github.com/stnava/ANTsR
}
Prior to installation of ANTsR, several external R packages
need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
\verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
\verb#rgl#, \verb#misc3d# which is facilitated by the 
\verb#install.packages()# mechanism.  Additionally, in order
to perform the supervised brain segmentation as described 
in later sections, one would need to also install 
\verb#randomForest#, \verb#snowfall#, \verb#rlecuyer#,
and \verb#ggplot2#.

CMake%
\footnote{
http://www.cmake.org/
}
is an open source tool for the management and building of 
large-scale software projects.  It is used
to coordinate the downloading of external packages,
such as the Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
}
and ANTs.  Detailed instructions for download and
installation can be found on the ANTsR github website.

\subsection{Supervised Brain Segmentation}

\subsubsection{Random Forests}

Building on the developments of Ho \cite{ho1995} and Amit and Geman \cite{amit1997},
Breiman introduced Random Forests for classification/regression \cite{breiman2001} which performs relatively well against other machine learning strategies such as support vector machines and neural networks.

\subsubsection{Multi-Modality Feature Image Preprocessing}

Although several studies have pointed out the importance of
intensity normalization and bias correction, our experience 
with the training data illustrated a degradation in performance
when one or both steps (using \cite{nyul2000} and \cite{tustison2010},
respectively) were performed due to the presence of the tumor/edema complex.  
Instead we simply windowed the image intensity
for all images to be between the quantiles $[0.01,0.99]$ and
subsequently rescaled to $[0,1]$.  This strategy seemed to provide 
the best classification (probably due to the fact that the intensity
corruption due to bias was relatively minimal in the data used).
However, improved results might be obtained
by a more sophisticated strategy in which an initial supervised segmentation 
is used to find the probabilistic estimates of the normal tissue
locations (as done in this work) which could then be followed by using
only the probabilistic maps of the normal tissues to construct a
weight, or confidence, mask for estimating the inhomogeneity bias
using N4 \cite{tustison2010} followed by a repeated application of 
the supervised segmentation strategy described.  Note that a similar
iterative-based strategy until convergence is implemented in the ANTs script \verb#antsAtroposN4.sh#.

\subsubsection{Multi-Modality Feature Image Generation}

Key to any supervised regression or classification protocol are the 
selected features for training and subsequent testing.  Based on previous
work and our own experience, we selected the following feature images
to showcase the supervised segmentation strategy developed in this work.

\begin{figure*}
  \centering
  \includegraphics[width=180mm]{Figures/featuresImages.pdf}
  \caption{Feature images from the BRATS\_HG0004 data set.}
\end{figure*}

\begin{itemize}
  \item Per modality (FLAIR, T1, T1C, T2)
    \begin{itemize}
      \item First-order neighborhood statistical images:
            mean, variance, skewness, and entropy. 
            Neighborhood radius = 3.
    \item GMM-based posteriors: CSF, gray matter, white matter, edema, and tumor
    \item GMM connected component geometry features:  volume, volume to 
          surface area ratio, eccentricity, elongation
    \item Template-based:  symmetric template difference and contralateral difference.
          Gaussian smoothing ($\sigma = 4$mm).
    \end{itemize}
  \item Misc.
    \begin{itemize}
    \item Normalized Euclidean distance
    \item Log Jacobian  
    \end{itemize}
\end{itemize}

For each modality, we create four first-order statistical feature images,
five Gaussian mixture model (GMM)-based posterior probability feature images,
four geometry features generated from the GMM posterior probability images
based on connected components, and two difference images using symmetric template
construction for a total of 4 modalities $\times$ (4 + 5 + 4 + 2) feature images per modality $=$ 60 images.  We employ two additional images consisting of the 
Euclidean distance image \cite{maurer2003} created from the skull-stripped 
binary mask rescaled to the range $[0,1]$ and the log
Jacobian image derived from the spatial normalization of the symmetric multivariate template and individual subject images.  Given the intensity corrected images
the corresponding multivariate template images, and a brain mask for each subject
creation of all feature images is performed using the script \verb#createFeatureImages.sh#.

Prior  cluster centers for specific tissue types learned from training data \cite{reynolds2009} are used in the GMM to create multiple feature images.  
Given $M$ tissue types (e.g. CSF, gray matter,
white matter, edema, and tumor), a GMM formulates the 
probability distribution at each voxel, $\mathbf{x}$, as the
sum of Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
\begin{align}
p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
\end{align}
where $\sum_{m=1}^M \lambda_m = 1$.  One popular method for 
determining the parameters of the GMM is maximum likelihood 
estimation which can performed using the Atropos segmentation 
tool \cite{avants2011}.  In contrast to previous generative
modeling approaches for multi-modal tumor segmentation 
(e.g. \cite{prastawa2003,zikic2012}), we do not use multivariate 
Gaussians to specify tissue probabilities but rather incorporate each
univariate probability map into the feature vector of the training
data.  As pointed out in \cite{menze2010}, multivariate modeling
might obscure the distinct biological information provided by each 
modality.  Instead, we let the random forest construction 
process determine the optimal combination of such multivariate
information.
Additionally, maximum posterior labeling from the GMM processing
is used to determine the connected components for each label.  
Geometric features (assigned voxel-wise) include the physical volumes 
of each connected component including the volume to surface area ratio, 
the elongation, and eccentricity.

In order to better characterize deviations from normal
multi-modal brain shape and appearance, several features were derived 
using population-specific multivariate template 
construction. A recent neuroimaging reproducibility study
by Landman et al. resulted in an open data cohort of 21
normal individuals, each imaged twice, comprising several
modalities including arterial spin labeling, 
fluid attenuated inversion recovery (FLAIR),
diffusion tensor imaging, functional imaging, T1, and T2 
\cite{landman2011}.

Given $K$ image modality types for $N$ subjects,  
${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
following cost function:
\begin{align}
  \sum_{n=1}^N 
        \Bigg[ D &\left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \\ \nonumber 
        +& \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg]
\end{align}
where $D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.  Further theoretical
details can be found in \cite{avants2008,avants2010}.
In terms of implementation, this algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a large computational cluster.

\begin{figure}
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=40mm]{Figures/S_templateFA_140.png} &
    \includegraphics[width=40mm]{Figures/S_templateFLAIR_140.png} \\
    (a) & (b) \\
    \includegraphics[width=40mm]{Figures/S_templateT1_140.png} &
    \includegraphics[width=40mm]{Figures/S_templateT2_140.png} \\
    (c) & (d) 
  \end{tabular}
  \caption{Multivariate symmetric template created from the Kirby 
           21 data described in \cite{landman2011}.  Shown are the
           (a) fractional anisotropy (FA), (b) FLAIR, (c) MPRAGE, 
           and (d) T2 template components.
          }
  \label{fig:symmetrictemplates}
\end{figure}

\subsubsection{\underline{Bra}in \underline{T}umor \underline{S}egmentation Challenge Data}

The Brain tumor segmentation
Associated with the 2012 International Conference on Medical Image Computing and Computer Assisted
Intervention (MICCAI),%
\footnote{
http://www.miccai2012.org
}
the

\subsubsection{MS lesion segmentation challenge Challenge Data}

\section{Results}

\begin{table*}
\caption{Dice scores from the MICCAI 2012 BRATs Study}
\begin{center}
\begin{tabular*}{0.975\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c}
\toprule
{} & \multicolumn{2}{c}{High-grade (real)} & \multicolumn{2}{c}{Low-grade (real)} & \multicolumn{2}{c}{High-grade (simulated)} & \multicolumn{2}{c}{Low-grade (simulated)}\\
{\bf Method} & Edema & Tumor & Edema & Tumor & Edema & Tumor & Edema & Tumor\\
\midrule
\cite{zikic2012} & {$0.70 \pm 0.09$} & {$0.71 \pm 0.24$} & {$0.44 \pm 0.18$} & {$0.62 \pm 0.27$} & {$0.65 \pm 0.27$} & {$0.90 \pm 0.05$} & {$0.55 \pm 0.23$} & {$0.71 \pm 0.20$} \\
\cite{bauer2012} & {$0.61 \pm 0.15$} & {$0.62 \pm 0.27$} & {$0.35 \pm 0.18$} & {$0.49 \pm 0.26$} & {$0.68 \pm 0.26$} & {$0.90 \pm 0.06$} & {$0.57 \pm 0.24$} & {$0.74 \pm 0.10$} \\
\rowcolor{yellow}
ANTsR & {$0.64 \pm 0.18$} & {$0.65 \pm 0.30$} & {$0.48 \pm 0.16$} & {$0.69 \pm 0.21$} & {$0.75 \pm 0.25$} & {$0.94 \pm 0.04$} & {$0.65 \pm 0.25$} & {$0.83 \pm 0.09$}  \\
\bottomrule
\end{tabular*}
\end{center}
\end{table*}

\section{Discussion and Conclusions}

@@

\section*{Acknowledgments}

>>>>

\title{Tilting at Algorithmic Windmills}

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{Brian B.~Avants (combined first author, second author or senior author?)}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, University of Pennsylvania,
                Philadelphia, PA}

@@

\begin{abstract} 
Exploration of neuroscience hypotheses have been greatly enhanced
by the increased availability of high-performance computational resources, large-scale
communal efforts such as the Insight Toolkit and the R statistical project and
algorithmic advancements for data transformations and processing. 
An integral component of the vetting process for novel algorithmic techniques
includes comparison with other methods previously
established within the community.  Public availability, including 
open source distribution, of established packages has significantly 
facilitated these comparative evaluations.  However, complementing the
recent set of papers pointing to serious methodological and statistical
bias considerations in neuroimaging research, we point out potential 
issues associated with a type of measurement bias in comparative algorithmic
evaluations and propose a set of guidelines for authors and reviewers
to minimize this confound.
\end{abstract}
\begin{keyword}
open science \sep reproducibility 
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

@@

%\section*{Acknowledgments}

\section{Notes/Thoughts}
\begin{itemize}
  \item Several kinds of bias have cropped up in the neuroimaging
  community recently:
  \begin{itemize}
    \item{methodological bias}:  Kreikesgorte
    \item{statistical bias}:  Vul2009, Tustison2012
    \item  What about the dead salmon fmri study---multiple comparisons correction?
    \item The single subject VBM study (assuming the single subject is representative of the population).
    \item{overview}:  9 circles of scientific hell ---neuroskeptic %http://blogs.discovermagazine.com/neuroskeptic/?p=1205#.UYZxyJWO7Tx
    \item Types of measurement bias: \cite{sackett1979}
    \begin{itemize}
      \item Instrument bias (code as a type of instrument which needs to be calibrated correctly).
      \item Expectation bias (i.e. confirmation bias)
    \end{itemize}  
  \end{itemize}
  We are concerned with the latter two.
  \item How does one avoid confirmation bias?  If I'm comparing my
  algorithm to somebody else's algorithm, I'm going to be naturally
  predisposed to confirmatory evidence that my algorithm is better
  while neglecting (most likely not with intent of doing so) evidence
  that disconfirms my original belief (that my algorithm is better).
  Richard Feynman quote (Cargo Cult Science):  
  \begin{quote} 
  The first principle [of science] is that you must not fool yourself—--and you are the easiest person to fool. So you have to be very careful about that. After you've not fooled yourself, it's easy not to fool other scientists. You just have to be honest in a conventional way after that.
  \end{quote}
  \item One particular aspect of the previous item is that it needs 
  to be highlighted if somebody codes up the algorithm
  to be compared themselves.  Writing bug-free code is difficult.  Look
  at how many bug fixes are provided by the ITK community on a daily
  basis.  
  \item Also, is simply saying that one algorithm is better than 
  another sufficient?  Should there be some theoretical explanation
  for it.  For example, when I wrote my DMFFD image registration paper,
  the comparison wasn't IRTK vs. DMFFD but rather FFD vs. DMFFD and
  there were theoretical reasons why performance should be better
  with the latter.  Also, I insisted to the reviewer on not using IRTK 
  for comparison since there are also so many other performance-related
  issues (type of interpolation, gradient step, metric, metric implementation, etc.) which would confound
  comparisons.  Also, it probably helped that the theme had a more specific,
  more verifiable focus, i.e. DMFFD produces a more efficient energy minimizer
  since it acts as a preconditioner in the standard gradient descent
  optimization.
  \item Comparisons are best made with publicly available data and
  results of the comparisons should be made available.
  \item Operating system needs to be defined.  Freesurfer results
  varied with MacOSx.
  \item Ideally, the authors of the new algorithm would work
  with the authors of the compared algorithm.  Given human
  nature, this type of cooperation is not guaranteed.  However,
  a minimum set of information is needed.
  \item Arno Klein as an example of somebody who did it right.  Takes a
  lot of work.
  \item 
  \item The source of the package software needs to be defined.  For 
  example, N4 has been
  instantiated in Slicer, ANTs, and c3d (also might be implemented
  in one of Styner's NITRC projects).  However, each might use
  different default parameters and have other tweaks which effects
  performance.  In Vladimir Fonov's github repository containing 
  various processing scripts for MINC, one can see from the history
  how N4 was used but then the users switched back to N3MNI (due to 
  performance issues?).  However, they used N4 out of the c3d package
  which the original authors (N.T./B.A.) haven't touched in three years
  so the parameters aren't optimal (shrink factors = 4, spline distance
  = 100 with 3 levels: $100\times50\times50$).  Changing these parameters 
  (which are crucial to performance) isn't
  accessible to the user from c3d.  
  \item Great resources for code-sharing
  	\begin{itemize}
		  \item github
		  \item NITRC
		  \item ITK
		\end{itemize}
\end{itemize}
